---
title: "Drought in IUCN Red List Assessments"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
date: "2024-05-16"
---

```{r, include=FALSE}
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

## Libraries required
```{r, message = F, warning = F}
library(tidyverse)
```

## Read in data
```{r}
droughtinRedListassessments <- read.csv("Data/Raw Data/Red List assessments with drought July 24.csv")
```

## Sourcing Functions
```{r}
source("Functions/Functions.R")
```

The data that we have here are all of the data downloaded from the IUCN Online Openly Accessible data portal that have Droughts (threat category 11.2) listed as a threat in their assessment. Interestingly, some don't actually list this in their key 'threat' paragraph in the IUCN (but at this point - July 2024 - I actually don't know if there's hidden information behind the assessments that you can download)

1425 species have drought listed. I can download assessment data for 1184 species

Now just want to check out how many actually list 'drought' specifically in their assessment

```{r}
#Drought specifically
assessmentswithdroughtlisted <- droughtinRedListassessments %>%
  filter(str_detect(threats, regex("drought", ignore_case = TRUE)))
# 657 species listed with 'drought'

#precipitation
assessmentswithprecipitationlisted <- droughtinRedListassessments %>%
  filter(str_detect(threats, regex("precipitation", ignore_case = TRUE)))
# 27 species listed with 'precipitation'

#rain
assessmentswithrainlisted <- droughtinRedListassessments %>%
  filter(str_detect(threats, regex("rain", ignore_case = TRUE)))
# 122 species listed with 'rain'
```




