---
title: "Drought Adaptive Capacity Analysis"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
date: "2024-05-16"
---


```{r, include=FALSE}
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output,
      sprintf('style="max-height: %s;"', options$max.height)
    )
    hook_output(x, options)
  })
})
```

## Libraries required

```{r, message = F, warning = F}
library(tidyverse)
library(ggplot2)
library(Hmisc)
library(ggridges)
library(emmeans)
library(ggbeeswarm)
```

## Read in data
```{r}
#read in the data and remove the aquatic species
droughtandxylemdata <- read.csv("Data/Data Used in Analyses/fulldroughtwithxylemdata.csv")
growthform <- read.csv("Data/Data Used in Analyses/fullIUCNandgrowthform.csv")

droughtandxylemdatafull <- left_join(droughtandxylemdata, growthform, by = 'scientificName') %>%
  filter(!Aquatic == "yes")
```

## Calculate safety margins of species based on Ψ50 - Ψmin

```{r}
droughtandxylemdatafull$safetymarginA <- as.numeric(droughtandxylemdatafull$psiminmidday) - as.numeric(droughtandxylemdatafull$p50)
droughtandxylemdatafull$safetymarginB <- as.numeric(droughtandxylemdatafull$psiminpredawn) - as.numeric(droughtandxylemdatafull$p50)
droughtandxylemdatafull$safetymargin <- rowMeans(droughtandxylemdatafull[,c("safetymarginA", "safetymarginB")], na.rm = T)
```

## Check overlap between robust drought data and p50 data
```{r}
#remove these species where there is no range at all in robust areas
droughtandxylemdatarobustp50 <- droughtandxylemdatafull %>%
  filter(EOODataAvailable == "AvailableEOO") %>%
  filter(!is.na(p50)) %>%
  filter(!(duration_245_proportion_non_robust == 1 &
      duration_585_proportion_non_robust == 1 &
      frequency_245_proportion_non_robust == 1 &
      frequency_585_proportion_non_robust == 1)) %>%
  mutate(threatened = if_else(.$redlistCategory.x %in% c('Critically Endangered','Endangered', 'Vulnerable'), "threatened", "ynot threatened"))

droughtandxylemdatarobustsafetymargin <- droughtandxylemdatafull %>%
  filter(EOODataAvailable == "AvailableEOO") %>%
  filter(!is.na(safetymargin)) %>%
  filter(!(duration_245_proportion_non_robust == 1 &
      duration_585_proportion_non_robust == 1 &
      frequency_245_proportion_non_robust == 1 &
      frequency_585_proportion_non_robust == 1)) %>%
  mutate(threatened = if_else(.$redlistCategory.x %in% c('Critically Endangered','Endangered', 'Vulnerable'), "threatened", "ynot threatened"))
```

## Check differences in p50 for threatened vs non-threatened species

### Regression model
```{r}
#run linear model
p50model <- lm(p50 ~ as.factor(threatened), data = droughtandxylemdatarobustp50)
#check assumptions
plot(p50model)
#get model output
summary(p50model)
```

### Plots
```{r}
p50threatenedgraph <- droughtandxylemdatarobustp50 %>%
  mutate(redlistCategory.x = if_else(.$redlistCategory.x %in% c('Critically Endangered','Endangered', 'Vulnerable'), "threatened", "ynot threatened")) %>%
  ggplot(aes(x = p50, y = threatened, fill = threatened)) +
  geom_density_ridges(scale = 0.7, alpha = 0.8, quantile_lines = T, quantile_fun = mean,
    jittered_points = TRUE,
    vline_size = 1, vline_color = "black",
    point_size = 3, point_alpha = 0.2, point_color = "black", 
    position = position_raincloud(adjust_vlines = FALSE, height = 0.18)) + 
  scale_y_discrete(expand = c(0, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0)) +   # for both axes to remove unneeded padding
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  scale_fill_manual(values = c("#E67F4B", "#7BCADE")) +
  guides(fill = "none") +
  xlab("P50") +
  ylab("") +
  theme_ridges() +
  geom_vline(xintercept = 0, size =1, color = "black", linetype = "dashed") +
  xlim(-10, 0)

p50threatenedgraph

#save the plot
ggsave(plot= p50threatenedgraph, filename = "Outputs/p50plot.jpeg",  device = "jpeg", width = 10, height = 9)
```

## Check differences in safety margins for threatened vs non-threatened species

### Regression model
```{r}
#run linear model
safetymarginmodel <- lm(safetymargin ~ as.factor(threatened), data = droughtandxylemdatarobustsafetymargin)
#check assumptions
plot(safetymarginmodel)
#get model output
summary(safetymarginmodel)
```

### Plot
```{r}
safetymarginthreatenedgraph <- droughtandxylemdatarobustsafetymargin %>%
  mutate(redlistCategory.x = if_else(.$redlistCategory.x %in% c('Critically Endangered','Endangered', 'Vulnerable'), "threatened", "ynot threatened")) %>%
  ggplot(aes(x = safetymargin, y = threatened, fill = threatened)) +
  geom_density_ridges(scale = 0.7, alpha = 0.8, quantile_lines = T, quantile_fun = mean,
    jittered_points = TRUE,
    vline_size = 1, vline_color = "black",
    point_size = 3, point_alpha = 0.2, point_color = "black", 
    position = position_raincloud(adjust_vlines = FALSE, height = 0.18)) + 
  scale_y_discrete(expand = c(0, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0)) +   # for both axes to remove unneeded padding
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  scale_fill_manual(values = c("#E67F4B", "#7BCADE")) +
  guides(fill = "none") +
  xlab("Safety Margin") +
  ylab("") +
  theme_ridges() +
  geom_vline(xintercept = 0, size =1, color = "black", linetype = "dashed") +
  xlim(-5, 6)

safetymarginthreatenedgraph

#save the plot
ggsave(plot= safetymarginthreatenedgraph, filename = "Outputs/safetymarginplot.jpeg",  device = "jpeg", width = 10, height = 9)
```

## Get means for results section for p50 and safety margins for threatened versus non-threatened species
```{r}
p50mean <- droughtandxylemdatarobustp50 %>%
  group_by(threatened) %>%
  summarise(meanp50 = mean(p50, na.rm = T))

safteymarginmean <- droughtandxylemdatarobustsafetymargin %>%
  group_by(threatened) %>%
  summarise(meansafetymargin = mean(safetymargin, na.rm = T))
```

