---
title: "Early Checks of Species lists"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
date: "2024-04-04"
---

## Libraries required

```{r}
library(tidyverse)
library(rredlist)
```

## Read in Data
```{r}
#data downloaded from the IUCN red list (https://www.iucnredlist.org/search) for all organisms in the "Plantae" Kingdom (~66,000 records) - plants
plantsonIUCN <- read.csv("Data/Raw Data/assessments.csv") %>%
  select(-language, -rationale, -threats, -population, -populationTrend, -range, -useTrade, -conservationActions, -realm, -yearLastSeen, -scopes)
taxonomyofplantsonIUCN <- read.csv("Data/Raw Data/taxonomy.csv")

#read in the data that we have spatial data for (and therefore drought data)
droughtdata <- read.csv("Data/Raw Data/drought_metrics_compile_20240326.csv") %>%
  rename(scientificName = species_name)
```

## Join data
```{r}
fullDroughtIUCNdata <- left_join(plantsonIUCN, droughtdata) %>%
  mutate(DroughtDataAvailable = ifelse(is.na(duration_min), 'NotAvailable', 'Available'))
```


## Some interesting checks of the data 

Quick look at how much our data represents
```{r}
howmuchdatawegot <- ggplot(data=fullDroughtIUCNdata, aes(DroughtDataAvailable, fill = DroughtDataAvailable)) +
  geom_bar() +
  scale_fill_manual(name = "DroughtDataAvailable", values = c("#7D5D3A", "#53619E")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot(howmuchdatawegot)
```


Aquatic plants (and species' "system")
```{r}
#Marine Species with Droughtdata
sum(fullDroughtIUCNdata$DroughtDataAvailable == "Available" & fullDroughtIUCNdata$systems == "Marine", na.rm=TRUE)
#122 species

#Freshwater species with drought data
sum(fullDroughtIUCNdata$DroughtDataAvailable == "Available" & fullDroughtIUCNdata$systems == "Freshwater (=Inland waters)", na.rm=TRUE)
#769! a lot...

#Terrestrial/Freshwater species with drought data
sum(fullDroughtIUCNdata$DroughtDataAvailable == "Available" & fullDroughtIUCNdata$systems == "Terrestrial|Freshwater (=Inland waters)", na.rm=TRUE)
#921! a lot...

#Terrestrial/marine species with drought data
sum(fullDroughtIUCNdata$DroughtDataAvailable == "Available" & fullDroughtIUCNdata$systems == "Terrestrial|Marine", na.rm=TRUE)
#50

#Terrestrial/Freshwater/Marine species with drought data
sum(fullDroughtIUCNdata$DroughtDataAvailable == "Available" & fullDroughtIUCNdata$systems == "Terrestrial|Freshwater (=Inland waters)|Marine", na.rm=TRUE)
#22
```

Quick bargraphs
```{r}
speciessystems <- ggplot(data=fullDroughtIUCNdata, aes(systems, fill = DroughtDataAvailable)) +
  geom_bar(position= "dodge") +
  scale_fill_manual(name = "DroughtDataAvailable", values = c("#7D5D3A", "#53619E")) +
  scale_y_log10() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot(speciessystems)
```


Aquatic plants problems: 
<br>
Problem 1) There are "freshwater" plants (769 species) that are included in the IUCN list AND our drought list - some of which ACTUALLY occupy water, some of which don't (grow along streams/rivers etc) and some of which can do both (juvenile in water and adult on land and vice versa) - what to do here as these plants may not actually be under "Drought" threat by the drought metrics we are using
<br>
Problem 2) Similar with "marine" plants (122 species) as above - but less are "both" or land dwelling near coast and more are actual sea grasses and other aquatic plants
<br>
Problem 3) Also related - there's these groups called "terrestrial/freshwater" (921 species) and "terrestrial/marine" (50 species) and all three (22 species) with a lot of species that have drought data! 
<br>

What to do for all of these aquatic plant cases - This is a total of 1884 species out of ~10,000... so a decent chunk. Do we look through one by one in their "habitat" description to select if they are aquatic or not? A lot of work! &#128128; &#128565; &#x1F635;




```{r}
plantsonIUCN
```






